---
title: "From a study to an analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From a study to an analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This walks through one analysis end to end: connecting, finding out what the
variables mean, fetching the data, joining two waves, and recording what a
published figure was computed from.

Nothing here runs when the vignette is built — the code needs a platform and a
token of your own.

## Connecting

Create a token under **your account menu → API Tokens** on your platform, then
put it in your environment rather than in the script:

```
# ~/.Renviron
SONDAVI_TOKEN=sdv_…
```

Restart R, and:

```{r}
library(sondavi)

con <- sondavi_connect("https://survey.your-institute.org")
```

A token written into an analysis script travels with it into version control,
onto shared drives and into supplementary material. That is the most common way
one leaks, which is why `sondavi_connect()` reads the environment by default.

## What is there

```{r}
sondavi_surveys(con)
#>    id                  title project    status       privacy response_count
#> 1  42       Wave 1 baseline   Ageing published     identified            318
#> 2  43       Wave 2 follow-up  Ageing published     identified            274
```

The token only ever shows the studies it was given. If one is missing, the token
was scoped to something narrower — that is the scope, not an error.

## What the variables mean

```{r}
cb <- sondavi_codebook(con, 42)
cb[, c("name", "label", "type")]
#>                  name                        label        type
#> 1                 age             How old are you?      string
#> 2                mood            How is your mood? categorical
#> 3 ratings.speed.score Rate the study — Speed × score     numeric
```

This is the part a CSV cannot carry: the intended type and the value labels
behind each code.

```{r}
cb$value_labels[[2]]
#>   value    text
#> 1     1     Bad
#> 2     2 Neutral
#> 3     3    Good
```

## The data

```{r}
d <- sondavi_responses(con, 42)
```

All pages are fetched for you, and the codebook is applied: `mood` arrives as a
factor with those labels rather than as `"1"`, `"2"`, `"3"`, and every column
carries its question text.

```{r}
table(d$mood)
attr(d$mood, "label")
#> [1] "How is your mood?"
```

### Two things worth knowing before you count anything

**Partial responses are included** when the study saves them, exactly as in the
platform's export. Someone who stopped halfway is a row whose `completed_at` is
`NA`:

```{r}
nrow(d)                                  # everyone who started
done <- subset(d, !is.na(completed_at))  # everyone who finished
```

**Timestamps are timestamps**, not text, so durations work directly:

```{r}
summary(as.numeric(difftime(done$completed_at, done$started_at, units = "mins")))
```

### Nested questions

Matrices and dynamic panels arrive as list columns — flattening them on arrival
would invent a shape the platform did not give. When you want the flat form:

```{r}
flat <- sondavi_unnest(d)
names(flat)
#> … "ratings.speed.score" "ratings.clarity.score"
```

Those are the same names the platform's own export writes, so an analysis built
on an exported file and one built on the API agree on what a variable is called.

## Joining waves

```{r}
d <- sondavi_waves(con, c(42, 43), names = c("baseline", "followup"))
```

Each wave's columns are suffixed, so `mood` becomes `mood_baseline` and
`mood_followup` instead of colliding. **Everyone seen in any wave is kept**: a
person who did not take the follow-up is still a row, with `NA` in its columns.

That is deliberate. Attrition is usually what a longitudinal design is about, and
an inner join would silently drop exactly the people you want to describe:

```{r}
table(returned = !is.na(d$mood_followup))
```

The match needs an identifier, so the token must carry the respondent ability and
the studies must not be anonymous. An anonymous study is not linkable by design,
and the function says so rather than guessing.

## Making the result citable

A live query is not a dataset. Run the same script next month and it may return
more rows, or fewer — new responses arrive, and retention or a data-subject
request may remove some.

The cheap version is to record the fingerprint next to your result:

```{r}
sondavi_fingerprint(d)
#> survey 42, 318 rows, fetched 2026-09-23T09:32:29+00:00, digest 332f3d2a1fbfd880
```

The better version, for anything you publish, is a snapshot:

```{r}
snap <- sondavi_snapshot(con, 42, label = "Paper, figure 2")
snap$id
#> [1] "8869174a-a6c7-4cbc-9a9e-2785dcecba51"
```

Put that UUID in the paper. A snapshot records *which* responses the figure was
computed from — never a copy of them, so erasure requests and retention keep
working on the real data.

Later, to reproduce:

```{r}
d <- sondavi_snapshot_responses(con, snap$id)
attr(d, "sondavi_complete")
#> [1] TRUE
```

If responses have been deleted since, you get a warning naming how many, and
`sondavi_complete` is `FALSE`. That is not an error — it is the answer you need
before re-running the analysis behind a published figure.

A colleague can replay your snapshot to reproduce your analysis. They never see
more than their own token allows: if yours carried identifiers and theirs does
not, they get the same rows with fewer columns, and are told that the recorded
fingerprint does not describe what they received.

## Fetching only what is new

For a study still collecting, ask for the increment rather than everything:

```{r}
recent <- sondavi_responses(con, 42, since = Sys.Date() - 7)
```

## When something goes wrong

The platform limits how often a token may be used and says how long to wait. The
package waits that long and retries; it does not hammer. This is the main reason
to use it rather than write the calls yourself — that, and every read being
recorded in the study's activity log under a recognisable user agent.
